The Chemistry Development Kit (CDK). 3. Atom typing, Rendering, Molecular Formula, and Substructure Searching

### 2016-07-16 article.tex
More content, e.g. about the rendering

% FIXME: finish migrating the author list

@@

  The new CDK version introduces a new API for stereochemistry information: the interfaces
  IStereoChemistryElement was introduced with the implementations TetrahedralChirality,
  for four coordinate compounds like bromo-chloro-fluoro-iodomethane,
  DoubleBondStereochemistry, for compounds like trans-2-butene, and
  ExtendedTetrahedral, for compounds like allene.

@@

  An implementation has been provided of the Signature structure descriptor for molecules 
  \cite{Faulon2003}. These act as a linear notation - like the SMILES format - 
  for the whole molecule as well as for connected substructures rooted at a single atom. The 
  descriptor can also be canonicalized to provide isomorphism-independent representations 
  \cite{Faulon2004}.

@@

@@

@@

The project continues to use PMD (http://pmd.sf.net/) for code quality checking,
but deviates from the default rules. For example, we are more liberal with 
variable name length. Moreover, a number of additional PMD tests have been
developed specifically for the CDK, that, for example, test if a class uses
the core interfaces instead of implementations of those interfaces. For example,
that the code uses IAtom instead of Atom. That said, these tests do generate a
few false positives, as the tests check the class name only, and not the
Java package the class is in.

@@

\begin{itemize}
\item \textbf{Project Name}: The Chemistry Development Kit
\item \textbf{Project home page}: \url{https://github.com/cdk/cdk}
\item \textbf{Operating system(s)}: Windows, GNU/Linux, OS/X
\item \textbf{Programming language}: Java
% FIXME: add appropriate references below
\item \textbf{Other (optional) requirements}: JNI-InChI, Vecmath, BEAM, Guava, JGraphT, Signatures, CMLDOM, XOM, JavaCC
\item \textbf{License}: LGPL v2.1 or later
\item \textbf{Any restrictions to use by non-academics}: None additional
\end{itemize}

@@

\section*{Figures}
  \subsection*{Figure 1 - Dependencies between CDK modules.}
  \label{fig:deps}
      Visualization of the dependencies between CDK modules. For example,
      the cdk-core depends on the cdk-interfaces module. A few higher level
      modules have been left out: cdk-builder3dtools, cdk-legacy, and
      cdk-depict.

@@

\subsection*{Figure 2 - Atom type information specified for a sp3-hybridized
carbon.}
\begin{verbatim}
  <at:AtomType rdf:ID="C.sp3">
    <at:categorizedAs rdf:resource="&cdkat;C.tetrahedral"/>
    <at:hasElement rdf:resource="&elem;C"/>
    <at:hybridization rdf:resource="&at;sp3"/>
    <at:formalCharge>0</at:formalCharge>
    <at:lonePairCount>0</at:lonePairCount>
    <at:formalBondType rdf:resource="&bo;single"/>
    <at:formalBondType rdf:resource="&bo;single"/>
    <at:formalBondType rdf:resource="&bo;single"/>
    <at:formalBondType rdf:resource="&bo;single"/>
  </at:AtomType>
\end{verbatim}

\subsection*{Figure 3 - The improved structure diagram generation has improved
code to solve overlap. The original SDG code used general heuristics (left) and the
OverlapResolver would fine tune the layout to ensure atoms would not be placed
at the same location (middle). The new SDG algorithm is able to
make more rigorous changes, making the final output must more pleasing (right).}

@@

@@

  \subsection*{Table 1 - A selection of key CDK modules with major changes.}
  \label{tab:modules}
  An overview of a selection of often used CDK modules with description,
  dependencies on third-party libraries, and the major changes since
  version~1.2. Depedencies between modules are depicted in Figure~\ref{fig:deps}.
  \vskip 1\baselineskip

    \begin{minipage}{1\textwidth}
    \renewcommand*{\thempfootnote}{\fnsymbol{mpfootnote}}
    \centering
    \begin{tabular}{lp{3cm}p{3cm}l}
  \textbf{Module}            & \textbf{Description}  & \textbf{Major Changes} & \textbf{Depedencies} \\ \hline
  interfaces                 &              & & Vecmath 1.5.2 \\ \hline
  core                       &              & & Google Guava 17.0 \\ \hline % FIXME: how to add https://github.com/google/guava ?
  standard                   &              & & \\ \hline
  render                     &              & Redesigned to make it more modular and support multiple widget toolkits, like AWT and SWT. & \\ \hline
  isomorphism                &              & & \\ \hline
  atomtype                   &              & Unified approach where atom typing is separated from other algorithms. & \\ \hline
  valencycheck               &              & & \\ \hline
  ioformats                  &              & & \\ \hline
  io                         &              & The MDL molfile reader has been rewritten and supports in atom types defind in the specification. & XPP3 1.1.4c \\ \hline
  iordf                      &              & New. & Jena 2.7.4 \\ \hline
  inchi                      &              & & JNI-InChI 0.8~\cite{Spjuth2013}  \\ \hline
  libiocml                   &              & & XOM 1.2.5, CMLXOM 3.1~\cite{Murray-Rust2011} \\ \hline
  smiles                     &              & SMILES support performance and converage is greatly improved. & Beam 0.9.1 \\ \hline %FIXME add a citation
  smarts                     &              & & Beam 0.9.1 \\ \hline %FIXME add a citation
  formula                    &              & New. & \\ \hline
  fingerprint                &              & Many new fingerprint types (see text). & Apache Commons Math 3.1.1 \\ \hline
  qsar                       &              & & XOM 1.2.5, JAMA 1.0.3~\cite{Hicklin2012} \\ \hline % FIXME: find JAMA reference
  signatures                 &              & & Signatures 1.1 \\ \hline % FIXME: ask Gilleain for the reference
  qsarmolecular              &              & & \\ \hline
  log4j                      &              & & Log4j 1.2.17 \\ \hline
    \end{tabular}
    \end{minipage}

>>>>

% FIXME: finish migrating the author list
% FIXME: add Gilleain

@@

  The new CDK version introduces a new API for stereochemistry information: the interfaces
  IStereoElement was introduced with the implementations TetrahedralChirality,
  for four coordinate compounds like bromo-chloro-fluoro-iodomethane,
  DoubleBondStereochemistry, for compounds like trans-2-butene, and
  ExtendedTetrahedral, for compounds like allene.

  An atom container has a method to iterate over all stereo elements:

\begin{verbatim}
for (IStereoElement element : container.) {
  if (element instanceof TetrahedralChirality) {
    // ...
  } else if (element instanceof DoubleBondStereochemistry) {
    // ...
  } else if (element instanceof ExtendedTetrahedral) {
    // ...
  }
}
\end{verbatim}

@@

  An implementation has been provided of the Signature structure descriptor for
  molecules~\cite{Faulon2003}. These act as a linear notation - like the SMILES format - 
  for the whole molecule as well as for connected substructures rooted at a single atom. The 
  descriptor can also be canonicalized to provide isomorphism-independent
  representations~\cite{Faulon2004}.

@@

  Moreover, a simplified API has been introduced that addresses most of the
  common rendering needs, with the DepictionGenerator class. To depict benzene
  the following code can be used:

\begin{verbatim}
new DepictionGenerator()
  .withSize(300, 300)
  .depict(benzene)
  .writeTo("BenzeneLocalized.png");
\end{verbatim}

  Many of the rendering options are available as parameters in the core API
  and as methods on the DepictionGenerator class. This includes subscructure
  coloring, exemplified with this example reaction (see Fig.~\ref{fig:depiction}):

\begin{verbatim}
TODO
\end{verbatim}

@@

However, perhaps the biggest factor in improved code quality is that we
instantiated a peer review process where any functionality changing patch is
required to be reviewed by one indepdent, senior CDK developer for  the development
branch, and two reviewers for stable branches. This patch development system
is supported by a number of automated validations steps as outlined below.

@@

The project continues to use PMD (url{http://pmd.sf.net/}) for code quality checking,
but deviates from the default rules. For example, we are more liberal with 
variable name length. Moreover, a number of additional PMD tests have been
developed specifically for the CDK, that, for example, test if a class uses
the core interfaces instead of implementations of those interfaces. For example,
that the code uses IAtom instead of Atom. That said, these tests do generate a
few false positives, as the tests check the class name only, and not the
Java package the class is in.

@@

\begin{itemize}
\item \textbf{Project Name}: The Chemistry Development Kit
\item \textbf{Project home page}: \url{https://github.com/cdk/cdk}
\item \textbf{Operating system(s)}: Windows, GNU/Linux, OS/X
\item \textbf{Programming language}: Java
% FIXME: add appropriate references below
\item \textbf{Other (optional) requirements}: JNI-InChI, Vecmath, Beam, Guava, JGraphT, Signatures, CMLDOM, XOM, JavaCC
\item \textbf{License}: LGPL v2.1 or later
\item \textbf{Any restrictions to use by non-academics}: None additional
\end{itemize}

@@

\section*{Figures}
  \subsection*{Figure 1 - Dependencies between CDK modules.}\label{fig:deps}
      Visualization of the dependencies between CDK modules. For example,
      the cdk-core depends on the cdk-interfaces module. A few higher level
      modules have been left out: cdk-builder3dtools, cdk-legacy, and
      cdk-depict.

@@

\subsection*{Figure 2 - Atom type information specified for a sp3-hybridized
carbon.}\label{fig:atomtype}
\begin{verbatim}
  <at:AtomType rdf:ID="C.sp3">
    <at:categorizedAs rdf:resource="&cdkat;C.tetrahedral"/>
    <at:hasElement rdf:resource="&elem;C"/>
    <at:hybridization rdf:resource="&at;sp3"/>
    <at:formalCharge>0</at:formalCharge>
    <at:lonePairCount>0</at:lonePairCount>
    <at:formalBondType rdf:resource="&bo;single"/>
    <at:formalBondType rdf:resource="&bo;single"/>
    <at:formalBondType rdf:resource="&bo;single"/>
    <at:formalBondType rdf:resource="&bo;single"/>
  </at:AtomType>
\end{verbatim}

\subsection*{Figure 3 - The improved structure diagram generation has improved
code to solve overlap. The original SDG code used general heuristics (left) and the
OverlapResolver would fine tune the layout to ensure atoms would not be placed
at the same location (middle). The new SDG algorithm is able to
make more rigorous changes, making the final output must more pleasing
(right).}\label{fig:sdg}

@@

\subsection*{Figure 4 - Integrated example showing the rendering and SMILES
parsing functionality.}\label{fig:depiction}

\includegraphics[width=\textwidth]{reaction.png}

@@

  \subsection*{Table 1 - A selection of key CDK modules with major changes.}\label{tab:modules}
  An overview of a selection of often used CDK modules with description,
  dependencies on third-party libraries, and the major changes since
  version~1.2. Depedencies between modules are depicted in Figure~\ref{fig:deps}.
  \vskip 1\baselineskip

    \begin{minipage}{1\textwidth}
    \renewcommand*{\thempfootnote}{\fnsymbol{mpfootnote}}
    \centering
    \begin{tabular}{lp{3cm}p{3cm}l}
  \textbf{Module}            & \textbf{Description}  & \textbf{Major Changes} & \textbf{Depedencies} \\ \hline
  interfaces                 &              & & Vecmath 1.5.2 \\ \hline
  core                       &              & & Google Guava 17.0 \\ \hline % FIXME: how to add https://github.com/google/guava ?
  standard                   &              & & \\ \hline
  render                     &              & Redesigned to make it more modular and support multiple widget toolkits, like AWT and SWT. & \\ \hline
  isomorphism                &              & & \\ \hline
  atomtype                   &              & Unified approach where atom typing is separated from other algorithms. & \\ \hline
  valencycheck               &              & & \\ \hline
  ioformats                  &              & & \\ \hline
  io                         &              & The MDL molfile reader has been rewritten and supports in atom types defind in the specification. & XPP3 1.1.4c \\ \hline
  iordf                      &              & New. & Jena 2.7.4 \\ \hline
  inchi                      &              & & JNI-InChI 0.8~\cite{Spjuth2013}  \\ \hline
  libiocml                   &              & & XOM 1.2.5, CMLXOM 3.1~\cite{Murray-Rust2011} \\ \hline
  smiles                     &              & SMILES support performance and converage is greatly improved. & Beam 0.9.1~\cite{beam} \\
  smarts                     &              & & Beam 0.9.1~\cite{beam} \\
  formula                    &              & New. & \\ \hline
  fingerprint                &              & Many new fingerprint types (see text). & Apache Commons Math 3.1.1 \\ \hline
  qsar                       &              & & XOM 1.2.5, JAMA 1.0.3~\cite{Hicklin2012} \\ \hline
  signatures                 &              & & Signatures 1.1 \\ \hline % FIXME: ask Gilleain for the reference
  qsarmolecular              &              & & \\ \hline
  log4j                      &              & & Log4j 1.2.17 \\ \hline
    \end{tabular}
    \end{minipage}
